y---
title: "detrendingdata"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(dplyr)
library(reshape)
library(ggthemes)
library(dplyr)
library(zoo)
library(ggplot2)
library(forecast)
library(mFilter)
library(forecast)
library(ggfortify)
library(changepoint)
```

```{r}
# daily flight data to reprensent national trends 
dailyflights <- statedata %>%
  group_by(date) %>%
  summarize(flightscount = sum(flightscount)) # total number of domestic flights per day 
dailyflights["rollingavg"] <- rollmean(dailyflights$flightscount, k = 7, fill = NA)
dailyflights["scaled"] <- scale(dailyflights$flightscount)
```

```{r}
# scale data for comparison
statedata$scaled <- scale(statedata$flightscount)
fl_all$scaled <- scale(fl_all$flightscount)
id_all$scaled <- scale(id_all$flightscount)
nj_all$scaled <- scale(nj_all$flightscount)
ny_all$scaled <- scale(ny_all$flightscount)
nm_all$scaled <- scale(nm_all$flightscount)
tx_all$scaled <- scale(tx_all$flightscount)

# vectors for each state
statev = c(dailyflights$flightscount)
flv= c(fl_all$scaled)
idv= c(id_all$scaled)
njv= c(nj_all$scaled)
nyv= c(ny_all$scaled)
nmv= c(nm_all$scaled)
txv= c(tx_all$scaled)
```

```{r}
# building time series 
statets <- ts(statev, start = c(2020,03,01), end = c(2021, 07, 31), frequency = 517)
flts <- ts(flv, start = c(2020,03,01), end = c(2021, 07, 31), frequency = 517)
idts <- ts(idv, start = c(2020,03,01), end = c(2021, 07, 31), frequency = 517)
njts <- ts(njv, start = c(2020,03,01), end = c(2021, 07, 31), frequency = 517)
nyts <- ts(nyv, start = c(2020,03,01), end = c(2021, 07, 31), frequency = 517)
nmts <- ts(nmv, start = c(2020,03,01), end = c(2021, 07, 31), frequency = 517)
txts <- ts(txv, start = c(2020,03,01), end = c(2021, 07, 31), frequency = 517)
```

## DETRENDING 

First, to detect the underlying trend, the time series is smoothed using a centered moving average. A moving window of 7 days represents weekly seasonality. 
```{r}
# STEP 1: DETECT THE TREND 

trend_states = ma(statets, order=7, centre =T) 
plot(as.ts(statets), ylab = "Flight Count", main = "Daily Flight Frequency Time Series") 
lines(trend_states, col = "red") 
legend(2020.77, 8000, legend = c("Smoothed Trend"), col = c("red"), lty = 1:1, cex = 0.8, box.lty = 0)
```

Next, removing the previously calculated trend from the time series will reveal seasonality 
```{r}
# STEP 2: DETREND THE TIME SERIES 

detrend = statets / trend_states
plot(as.ts(detrend), ylab = "Flight Count", main = "Detrended Time Series - All Flights")
```

From the detrended time series, the average seasonality is computed by adding the seasonality together and dividing by the seasonality period. To average together the time series it is fed into a matrix which is then transformed to contain elements of the same period (like day of week). Finally, the mean of each column in computed. 
```{r}
# STEP 3: AVERAGE THE SEASONALITY

m_air = t(matrix(data = detrend, nrow = 74))
seasonal_allflights = colMeans(m_air, na.rm = T)
plot(as.ts(rep(seasonal_allflights,74)))
```
The multiplicative formula is Time Series = Seasonal * Trend * Random. Plotting the result will leave behind the random noise of the time series. 
```{r}
# STEP 4: EXAMINING REMAINING RANDOM NOISE 

random_allflights = statets / (trend_allflights * seasonal_allflights)
plot(as.ts(random_allflights))

# detrending the individual states using national trends 
random_fl = flts / (trend_allflights * seasonal_allflights)
random_id = idts / (trend_allflights * seasonal_allflights)
random_nj = njts / (trend_allflights * seasonal_allflights)
random_nm = nmts / (trend_allflights * seasonal_allflights)
random_ny = nyts / (trend_allflights * seasonal_allflights)
random_tx = txts / (trend_allflights * seasonal_allflights)
```

```{r}
# visualizing random noise for each state 
plot(random_fl, main = "Florida Random Noise")
plot(random_id, main = "Idaho Random Noise")
plot(random_nj, main = "New Jersey Random Noise")
plot(random_nm, main = "New Mexico Random Noise")
plot(random_ny, main = "New York Random Noise")
plot(random_tx, main = "Texas Random Noise")
```

A decomposition of the time series visualizes all of the above steps together. 
```{r}
ts_flights = ts(statets, frequency = 52)
decompose_flights = decompose(ts_flights, "multiplicative")

plot(as.ts(decompose_flights$seasonal))
plot(as.ts(decompose_flights$trend))
plot(as.ts(decompose_flights$random))
plot(decompose_flights)
```
  
# DETECTING CHANGE POINTS 
```{r}
# FLORIDA

flcp <- cpt.var(na.omit(random_fl), method = "PELT")
plot(flcp, type = "l", cpt.col = "red", xlab = "Date", ylab = "", yaxt="none", cpt.width = 4, main = "Florida Changepoints")
cpts(flcp) # output is the specific changepoints identified by the algorithm 

fl_all %>%
  select(date, flightscount) %>%
  mutate(growth = 100 * ((flightscount / lag(flightscount)) - 1)) %>%
  filter(date > "1960-01-01") %>%
  slice(cpts(flcp))
```

```{r}
# IDAHO

idcp <- cpt.var(na.omit(random_id), method = "PELT")
plot(idcp, type = "l", cpt.col = "red", xlab = "Date", ylab = "", yaxt="none", cpt.width = 4, main = "Idaho Changepoints")
cpts(idcp)

id_all %>%
  select(date, flightscount) %>%
  mutate(growth = 100 * ((flightscount / lag(flightscount)) - 1)) %>%
  filter(date > "1960-01-01") %>%
  slice(cpts(idcp))
```

```{r}
# NEW MEXICO

nmcp <- cpt.var(na.omit(random_nm), method = "PELT")
plot(nmcp, type = "l", cpt.col = "red", xlab = "Date", ylab = "", yaxt="none", cpt.width = 4, main = "New Mexico Changepoints")
cpts(nmcp)

nm_all %>%
  select(date, flightscount) %>%
  mutate(growth = 100 * ((flightscount / lag(flightscount)) - 1)) %>%
  filter(date > "1960-01-01") %>%
  slice(cpts(nmcp))
```

```{r}
# NEW YORK

nycp <- cpt.var(na.omit(random_ny), method = "PELT")
plot(nycp, type = "l", cpt.col = "red",xlab = "Date", ylab = "", yaxt="none", cpt.width = 4, main = "New York Changepoints")
cpts(nycp)

ny_all %>%
  select(date, flightscount) %>%
  mutate(growth = 100 * ((flightscount / lag(flightscount)) - 1)) %>%
  filter(date > "1960-01-01") %>%
  slice(cpts(nycp))
```

```{r}
# NEW JERSEY

njcp <- cpt.var(na.omit(random_nj), method = "PELT")
plot(njcp, type = "l", cpt.col = "red", xlab = "Date", ylab = "", yaxt="none",cpt.width = 4, main = "New Jersey Changepoints")
title(xlab = "Date", ylab = "")
cpts(njcp)

nj_all %>%
  select(date, flightscount) %>%
  mutate(growth = 100 * ((flightscount / lag(flightscount)) - 1)) %>%
  filter(date > "1960-01-01") %>%
  slice(cpts(njcp))
```

```{r}
# TEXAS

txcp <- cpt.var(na.omit(random_tx), method = "PELT")
plot(txcp, type = "l", cpt.col = "red", cpt.width = 4, 
     xlab = "Date", ylab = "", yaxt="none", main = "Texas Changepoints")
abline(v = as.Date("2020-03-26"))
cpts(txcp)

tx_all %>%
  select(date, flightscount) %>%
  mutate(growth = 100 * ((flightscount / lag(flightscount)) - 1)) %>%
  filter(date > "1960-01-01") %>%
  slice(cpts(txcp))
```
